Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-03 04-analisis.qmd
update interaction effects

@@

# interactions

m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc + merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

@@

@@

@@

m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent + ola_num| idencuesta),
                data = df_study1)

@@

@@

@@

texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.79,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::texreg(list(m8,m9),
               custom.model.names = c("Model 8", "Model 9"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("Yes",2))))

@@

@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences. In neither the within-person specification (Model 8) nor the between-person specification (Model 9) do the interaction terms reach statistical significance, indicating that higher perceptions of economic inequality do not meaningfully alter the effect of meritocracy on support for market allocation of social services.

@@

p4 <- ggpredict(m11, terms = c("ola_num [all]", "merit_effort")) %>% 
  plot() + 
  scale_x_continuous(n.breaks = 6) +
  labs(x = "Wave",
       title = "(A) People are rewarded for their efforts",
       color = NULL) +
  ggdist::theme_ggdist() +
  theme(legend.position = "none")

p5 <- ggpredict(m12, terms = c("ola_num [all]", "merit_talent")) %>% 
  plot() + 
  scale_x_continuous(n.breaks = 6) +
  labs(x = "Wave",
       title = "(B) People are rewarded for their intelligence",
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)",
       color = NULL) +
  ggdist::theme_ggdist() +
  theme(legend.position = "bottom")

p4 / p5

@@

With respect to the temporal dynamics of the key predictors, @fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences (complete models appear in the Supplementary Material). As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates. Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

>>>>

df_study1 <- df_study1 %>%
  mutate(
    perc_inequality_cwc_dic = ifelse(perc_inequality_cwc > mean(perc_inequality_cwc, na.rm = TRUE), "mayor", "menor"),
    merit_effort_cwc_dic = ifelse(merit_effort_cwc > mean(merit_effort_cwc, na.rm = TRUE), "mayor", "menor"),
    merit_talent_cwc_dic = ifelse(merit_talent_cwc > mean(merit_talent_cwc, na.rm = TRUE), "mayor", "menor"),
    perc_inequality_mean_dic = ifelse(perc_inequality_mean > mean(perc_inequality_mean, na.rm = TRUE), "mayor", "menor"),
    merit_effort_mean_dic = ifelse(merit_effort_mean > mean(merit_effort_mean, na.rm = TRUE), "mayor", "menor"),
    merit_talent_mean_dic = ifelse(merit_talent_mean > mean(merit_talent_mean, na.rm = TRUE), "mayor", "menor")
  ) %>%
  mutate(across(ends_with("_dic"), factor, levels = c("menor", "mayor")))

@@

# interactions 
m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc_dic*perc_inequality_cwc_dic + merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

m8.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc +
              (1 + merit_effort_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

m8.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_cwc*perc_inequality_cwc +
              (1 + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                data = df_study1)

@@

m9.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1)

m9.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_talent_mean*perc_inequality_mean +
              (1 | idencuesta),
                data = df_study1)

@@

m10.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola_num + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
                data = df_study1)

m10.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola_num + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
                data = df_study1)

@@

m11.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola_num  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1)

m11.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola_num + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
                data = df_study1)

m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent + ola_num| idencuesta),
                data = df_study1)

m12.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1)

m12.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean*ola_num + educ + quintil1 + ess + ideo + sex + age + 
              (1 + merit_talent_mean + ola_num| idencuesta),
                data = df_study1)

@@

m13.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             merit_effort_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
                data = df_study1)

m13.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
             ola_num*merit_effort_mean*perc_inequality_mean +
              (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
                data = df_study1)

@@

m14.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_cwc*perc_inequality_cwc*ola_num +
              (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
                data = df_study1)

m14.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
              merit_talent_mean*perc_inequality_mean*ola_num +
              (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
                data = df_study1)

#knitreg(list(m13.1, m13.2,m14.1, m14.2))
```

```{r}
interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
```

@@

texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.79,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))

@@

```{r}
#| label: fig-interact
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Within effects of perception of inequality and meritocratic perceptions on market justice preferences"
#| fig-cap-location: top
#| echo: false
#| results: asis

p6 <- ggpredict(m9.1, terms = c("merit_effort_mean", "perc_inequality_mean")) %>% 
  plot() + 
 # scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Perception\nInequality (BE)",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")) +
  labs(x = "Merit: Effort (BE)",
       title = "(A) People are rewarded for their efforts",
       color = NULL) +
  ggdist::theme_ggdist() +
  theme(legend.position = "bottom")

p7 <- ggpredict(m9.2, terms = c("merit_talent_mean", "perc_inequality_mean")) %>% 
  plot() +
#  scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Perception\nInequality (BE)",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  labs(
    x = "Merit: Talent (BE)",
    title = "(B) People are rewarded for their intelligence",
    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
  ) +
  ggdist::theme_ggdist() +
  theme(legend.position = "bottom")

ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)

@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences. The interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance. However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services. These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater. These interaction effects are visualized in @fig-interact

```{r echo=FALSE, results='asis'}
#| label: tbl-interactions2
#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
#| tbl-cap-location: top

ccoef <- list(
  "(Intercept)" = "Intercept",
  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
    )

texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("Yes",6))))

@@

p4 <- ggpredict(m11.2, terms = c("ola_num [all]", "merit_effort_mean")) %>% 
  plot() + 
  scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Merit:\nEffort",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "Merit:\nEffort",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")) +
  labs(x = "Wave",
       title = "(A) People are rewarded for their efforts",
       color = NULL) +
  ggdist::theme_ggdist() +
  theme(legend.position = "top")

p5 <- ggpredict(m12.2, terms = c("ola_num [all]", "merit_talent_mean")) %>% 
  plot() +
  scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Merit:\nTalent",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "Merit:\nTalent",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  labs(
    x = "Wave",
    title = "(B) People are rewarded for their intelligence",
    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
  ) +
  ggdist::theme_ggdist() +
  theme(legend.position = "top")

ggpubr::ggarrange(p4, p5, common.legend=FALSE, ncol=1)

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables (Model 16 and Model 17). These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated. @fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences. As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates. Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.
